---
title: "homework_3"
output: html_document
date: "2025-04-11"
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

```{r init}
library(dplyr)
library(Seurat)
library(patchwork)
#for an error: object ‘markvario’ is not exported by 'namespace:spatstat'
remotes::install_version("spatstat", version = "1.64-1")
```
# Homework Problem 1
Analyze the 68k PBMCs dataset in the same way as presented in the Seurat's guide with PBMC3k. Apply QC, PCA, jackstraw, clustering, and t-SNE to create figure similar to Figure 3b on Zheng et al. 2017. Note that there are differences between Zheng's original analysis and Seurat's analysis. Pay attentions to hyper-parameters that you must choose for this new bigger dataset.

```{r}
# Load the 68k PBMC dataset
pbmc68k.data <- Read10X(data.dir = "C:/Users/zmzuz/import_zuza/studia_II/1_rok/2_semestr/modelowanie/lab6/fresh_68k_pbmc_donor_a/filtered_matrices_mex/hg19")
pbmc68k <- CreateSeuratObject(counts = pbmc68k.data, project = "pbmc68k", 
                             min.cells = 3, min.features = 200)
pbmc68k
```

```{r}
head(pbmc68k@meta.data, 5)
```

```{r}
pbmc68k.data[c("CD3D","TCL1A","MS4A1"), 1:30]
```
Quality Control
```{r}
# mitochondrial gene percentage
pbmc68k[["percent.mt"]] <- PercentageFeatureSet(pbmc68k, pattern = "^MT-")
```
```{r}
# Visualize QC metrics
VlnPlot(pbmc68k, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(pbmc68k, feature1 = "nCount_RNA", feature2 = "percent.mt") 
plot2 <- FeatureScatter(pbmc68k, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") 
plot1 + plot2

pbmc68k <- subset(pbmc68k, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

```
Normalization and Feature Selection
```{r}
pbmc68k <- NormalizeData(pbmc68k)
pbmc68k <- FindVariableFeatures(pbmc68k, selection.method = "vst", nfeatures = 2000)
```

```{r var_features, fig.height=5, fig.width=11}

# 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc68k), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc68k)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```
Scaling and PCA
```{r}
all.genes <- rownames(pbmc68k)
pbmc68k <- ScaleData(pbmc68k, features = all.genes)
pbmc68k <- RunPCA(pbmc68k, features = VariableFeatures(object = pbmc68k))
```
Determine Dimensionality
```{r pca_viz, message=TRUE}

print(pbmc68k[['pca']], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc68k, dims = 1:2, reduction = 'pca')
DimPlot(pbmc68k, reduction = 'pca')
```

```{r single-heatmap}
DimHeatmap(pbmc68k, dims = 1, cells = 500, balanced = TRUE)
```

```{r multi-heatmap, fig.height=15, fig.width=9}
DimHeatmap(pbmc68k, dims = 1:15, cells = 500, balanced = TRUE)
```


```{r}
# JackStraw procedure (commented out for speed)
#pbmc68k <- JackStraw(pbmc68k, num.replicate = 100)
#pbmc68k <- ScoreJackStraw(pbmc68k, dims = 1:20)

JackStrawPlot(pbmc68k, dims = 1:20)
```



```{r}
# Elbow plot
ElbowPlot(pbmc68k, ndims = 50)
```
Clustering
```{r}
pbmc68k <- FindNeighbors(pbmc68k, dims = 1:15)
pbmc68k <- FindClusters(pbmc68k, resolution = 0.8)  # Increased resolution for larger dataset
```
```{r}
head(Idents(pbmc68k), 5)
```

t-SNE Visualization
```{r}
pbmc68k <- RunTSNE(pbmc68k, dims = 1:15)
DimPlot(pbmc68k, reduction = "tsne", label = TRUE)
```

```{r saveobject, eval=FALSE}
dir.create("../output", showWarnings = FALSE, recursive = TRUE)

saveRDS(pbmc68k, file = "../output/pbmc68k_tutorial.rds")
```
Differential Expression Analysis
```{r}
# markers for all clusters
pbmc68k.markers <- FindAllMarkers(pbmc68k, only.pos = TRUE, 
                                 min.pct = 0.25, logfc.threshold = 0.25)

# Top markers per cluster
pbmc68k.markers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_log2FC)
```

Visualization of marker genes 
```{r}
# Visualize marker genes
FeaturePlot(pbmc68k, 
            features = c("CCR7", "LEF1", "GZMK", "DUSP2", "IL7R", "S100B"),
            reduction = "tsne",  
            ncol = 3,            
            order = TRUE,        
            pt.size = 0.5,       
            cols = c("lightgrey", "blue"), 
            combine = TRUE) +    
  plot_annotation(title = "Marker genes")
```

Cluster annotation
```{r}
cluster_levels <- levels(pbmc68k)
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", 
                    "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet",
                    "Activated CD4 T", "Progenitor", "Other",
                    rep("Unknown", length(cluster_levels) - 12))
names(new.cluster.ids) <- cluster_levels
pbmc68k <- RenameIdents(pbmc68k, new.cluster.ids)
DimPlot(pbmc68k, reduction = "tsne", label = TRUE, pt.size = 0.5) + NoLegend()
```

differences from the PBMC3k tutorial include:
-Adjusted QC thresholds (higher nFeature cutoff for larger dataset)
-Used more PCs (15 instead of 10) to capture additional variation
-Increased clustering resolution (0.8 instead of 0.5) to detect more subpopulations
-Added t-SNE visualization instead of UMAP for better comparison with Zheng et al.

# Homework Problem 2
Create a hierachical clustering by applying K-means clustering to cells defined by each of 10 cluster. Try to find a suitable number of clusters (k) for each sub-population.

Present

* overall t-SNE visualization with all clusters -- make sure that hierarchy (group) is visualized. 
* visualize t-SNE for each of 10 clusters, with their sub-clusters.

```{r}

pbmc68k <- readRDS("C:/Users/zmzuz/import_zuza/studia_II/1_rok/2_semestr/modelowanie/output/pbmc68k_tutorial.rds")

Idents(pbmc68k) <- "seurat_clusters"
```


```{r}
# hierarchical clustering
pbmc68k$main_cluster <- Idents(pbmc68k) 
pbmc68k$subcluster <- as.character(Idents(pbmc68k))  

# Function to determine optimal k ( elbow method)
find_optimal_k <- function(data, max_k = 5) {
  wss <- sapply(1:max_k, function(k) {
    kmeans(data, centers = k, nstart = 25)$tot.withinss
  })
  which.min(diff(wss)/wss[-length(wss)]) + 1
}
```

```{r}
# Perform hierarchical clustering for each main cluster
pbmc68k <- FindClusters(pbmc68k, resolution = 0.9)


pbmc68k$hierarchical <- NA
for (cluster in 0:9) {
  if (cluster %in% levels(Idents(pbmc68k))) {
    cells <- WhichCells(pbmc68k, idents = cluster)
    subset_data <- subset(pbmc68k, cells = cells)
  
  if (!"pca" %in% names(subset_data@reductions)) {
    subset_data <- RunPCA(subset_data, npcs = 15)
  }
}
    
  
  # PCA embeddings
  pca_data <- Embeddings(subset_data, "pca")[,1:15]
  
  # Determine optimal k 
  optimal_k <- max(2, find_optimal_k(pca_data))
  
  # Perform K-means clustering
  km_res <- kmeans(pca_data, centers = optimal_k, nstart = 25)
  
  # Assign subcluster IDs
  pbmc68k$subcluster[cells] <- paste0(cluster, "_", km_res$cluster)
}
```


```{r plots, fig.width=10, fig.height=4}
# Visualization
library(ggplot2)
# color palette for hierarchical visualization
main_colors <- scales::hue_pal()(length(levels(Idents(pbmc68k))))
subcluster_colors <- unlist(lapply(1:length(main_colors), function(i) {
  shades <- colorRampPalette(c(main_colors[i], "white"))(7)[2:4]
  shades[1:3] 
}))

# t-SNE with hierarchy
p1 <- DimPlot(pbmc68k, group.by = "main_cluster", label = TRUE) + 
  ggtitle("Main Clusters (10 groups)")
p2 <- DimPlot(pbmc68k, group.by = "subcluster", label = FALSE, cols = subcluster_colors) +
  ggtitle("Hierarchical Subclusters")
combined_plot <- p1 + p2
combined_plot

# Individual cluster visualizations
subcluster_plots <- list()
for (cluster in levels(Idents(pbmc68k))) {
  cells <- WhichCells(pbmc68k, idents = cluster)
  subset_data <- subset(pbmc68k, cells = cells)
  
  subcluster_plots[[cluster]] <- DimPlot(subset_data, 
                                       group.by = "subcluster",
                                       label = TRUE) +
    ggtitle(paste("Cluster", cluster, "Subclusters")) +
    theme(legend.position = "none")
}

```
```{r}
head(pbmc68k@meta.data, 5)
```


```{r plots, fig.width=10, fig.height=12 }
# all subcluster plots
wrap_plots(subcluster_plots, ncol = 3)
```

```{r}
# Save results
saveRDS(pbmc68k, "pbmc68k_hierarchical.rds")
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
