
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Analysis for amygdala exc. an hipp. ca1-3
#TSNE amygdala exc. scz2022
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)


# 1. tsne data
dat.tsne <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy")
names(dat.tsne)
# 2. sldsc data
dat.cluster.res.amy <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Amygdala excitatory")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res.amy, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 0.75,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}


output_path <- file.path(getwd(), "..", "figures", "tSNE_amygdala_excitatory_scz2022.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
#SNP-heritability enrichment by cluster- AMygdala exc- data from paper

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep

dat.cluster.res.amy <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Amygdala excitatory")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 2. plot
p <- ggplot(dat.cluster.res.amy, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Amygdala Excitatory",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_amygdala_excitatory_enrichment_sorted.pdf",
       plot = p,
       width = 10, 
       height = 6)

```
#SNP-heritability enrichment by cluster- AMygdala exc- my sldsc data

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep
dat.cluster.res.my <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy\n")
names(dat.tsne)
df <- fread("../data/amygdala_sldsc.csv")

# 3. Function Fisher 
combine_pvalues_fisher <- function(p) {
  chisq_stat <- -2 * sum(log(p), na.rm = TRUE)
  df <- 2 * sum(!is.na(p))
  p_combined <- pchisq(chisq_stat, df = df, lower.tail = FALSE)
  return(p_combined)
}

# 4. Agregation p-value (Fisher)
dat.cluster.res.my <- df %>%
  group_by(Cluster_num) %>%
  summarise(
    combined_pvalue = combine_pvalues_fisher(Enrichment_p),
    minuslog10P = -log10(combined_pvalue),
    if.sig.fdr = ifelse(combined_pvalue < 0.05, "yes", "no")
  ) %>%
  # Sort by minuslog10P in descending order and convert to factor
  arrange(desc(minuslog10P)) %>%
  mutate(Cluster_num = factor(Cluster_num, levels = Cluster_num))

# 2. plot
p <- ggplot(dat.cluster.res.my, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         midpoint = 0.4,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Amygdala Excitatory",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_amygdala_excitatory_enrichment_sorted_my.pdf",
       plot = p,
       width = 10, 
       height = 6)

```
tsne plot scz2022- AMygdala exc- data from paper
```{r}
# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res.amy, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_amygdala_excitatory_scz2022_sup.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
TSNE plot  Fisher combined p-values for AMygdala - my sldsc data
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(scico)

# 1. load data
dat.tsne <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy\n")
names(dat.tsne)
df <- fread("../data/amygdala_sldsc.csv")

# 3. Function Fisher 
combine_pvalues_fisher <- function(p) {
  chisq_stat <- -2 * sum(log(p), na.rm = TRUE)
  df <- 2 * sum(!is.na(p))
  p_combined <- pchisq(chisq_stat, df = df, lower.tail = FALSE)
  return(p_combined)
}

# 4. Agregation p-value (Fisher)
dat.cluster.res <- df %>%
  group_by(Cluster_num) %>%
  summarise(
    combined_pvalue = combine_pvalues_fisher(Enrichment_p),
    minuslog10P = -log10(combined_pvalue),
    if.sig.fdr = ifelse(combined_pvalue < 0.05, "yes", "no")
  ) %>%
  mutate(Cluster_num = as.character(Cluster_num))

cat("sldsc data\n")
print(dat.cluster.res)

# 5. Merge  tsne
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%
  left_join(dat.cluster.res, by = "Cluster_num")  

# 6. Labels 
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 7. Plot TSNE
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "-log10(P) (Fisher combined)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5", size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# 8. adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}

# 9. file save
output_path <- file.path(getwd(), "..", "figures", "tsne_amygdala_sldsc_fisher.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)


```
tsne plot for hippocampal ca1-3- data from paper
```{r}
# 1. tsne data
dat.tsne <- fread("../data/hipp_tsne.csv.gz")

dat.cluster.res <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Hippocampal CA1-3")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 
# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_hippocampal_ca1-3_scz2022.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
#SNP-heritability enrichment by cluster- hippocampal ca1-3 - data from paper
```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep
dat.cluster.res.sup <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Hippocampal CA1-3")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 2. plot
p <- ggplot(dat.cluster.res.sup, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Hippocampal CA1-3",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_hippocampal_enrichment_sorted.pdf",
       plot = p,
       width = 10, 
       height = 6)

```
