---
title: "Enitre analysis"
output: html_document
date: "2025-06-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

analysis from paper
# fig_2a.R
#- Fig2A: Superclusters enrichment significance of traits' SNP-heritability
```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(readxl)
library(scico)
source("../scripts/order_data.R"))

dat1 <- read_xlsx("../data/Supplementary_Datasets.xlsx", sheet = "Supplementary_Data_3") %>%
  rename(Trait=label)

dat1$Supercluster <- factor(dat1$Supercluster, 
                            levels=rev(order_ct))
dat1$Trait <- factor(dat1$Trait, 
                    levels=order_final_traits)
#- bubble plot
p <- ggplot(dat1,
            aes(x=Trait, y=Supercluster,
                fill = -log10(P), 
                size = if.sig.fdr,
                color = if.sig.fdr)) + 
  geom_point(alpha = .85, shape = 21) +
  theme(panel.background = element_rect(fill = "transparent"),
        axis.text.x=element_text(angle = 30, hjust=1),
        legend.key = element_rect(fill = "transparent")) +
  scico::scale_fill_scico(palette = "lajolla",direction=-1, midpoint = 5,
                          name=expression('-log'[10]*'(P)')) + 
  facet_grid(~fct_relevel(disorder_type,
                          "psychiatric",
                          "brain traits",
                          
                          ), 
             scales = "free_x", 
             space = "free_x") +
  scale_size_manual(values=c(1.5, 5))+
  scale_color_manual(values=c("no"="grey","yes"="black")) + 
  labs(size = "Sig.(FDR)")+ ylab("") +xlab("") +
  guides(color="none") + 
  guides(size=guide_legend(override.aes=list(color=c("lightgrey","black")))) 

#- bar plot
dat2 <- read_xlsx("../data/Supplementary_Datasets.xlsx",sheet = "Supplementary_Data_1") 
dat2$label = factor(dat2$label, 
                    levels=order_final_traits)
p1 <- ggplot(dat2 %>% filter(label%in%dat1$Trait),
             aes(x=label, y=ldsc_h2)) +
  geom_bar(stat="identity",width = 1, color="white")+ 
  facet_grid(~fct_relevel(disorder_type,
                          "psychiatric",
                          "brain traits",
                          ), 
             scales = "free_x", 
             space = "free_x") +
  theme(panel.background = element_rect(fill = "transparent"),
        axis.text.x=element_text(angle = 30, hjust=1),
        axis.ticks.x = element_blank(),
        legend.key = element_rect(fill = "transparent")) +
  xlab("")


#- output
ggsave(p,
       filename = "../figures/2a_bubble.pdf",
       width = 16, 
       height = 7.5)
ggsave(p1 + theme(axis.text.x=element_blank()),
       filename = "../figures/2a_bar.pdf",
       width = 16, 
       height = 1.5)
```

# fig_3a.R
# Fig3A: tSNE of neuronal cells, colored by cluster-level significance of schizophrenia (scz2022) SNP-heritability enrichment.
#-- Using **large loom and tSNE files** from Siletti et al, Science 2023 paper; download link provided below
#-- loom file (adult_human_20221007.loom): can be downloaded from https://storage.cloud.google.com/linnarsson-lab-human/adult_human_20221007.loom
#-- t-SNE coordinates (neuron_tsne.csv.gz): from the authors of the paper Siletti et al, Science 2023. Provided at the current repository under: /workflow/neuron_tsne.csv.gz

```{r}
library(data.table)
library(tidyverse)
library(loomR)
library(here)
library(ggplot2)

# 1. Get cell ID and cluster from the loom file
lfile <- connect(filename = "../data/adult_human_20221007.loom", #-- **User need to download the loom file and specify the path here** --#
                 mode = "r", skip.validate = TRUE)
lfile
names(lfile[["col.attrs"]])
CellID <- lfile[["col_attrs/CellID"]][]
Cluster <- lfile[["col_attrs/Clusters"]][] # used to link to Supercluster
dat <- cbind(CellID, Cluster) %>% as.data.frame()

# Link to Superclusters  
library(readxl)
CT_annot <- read_excel("../data/silleti_table_s2.xlsx") %>%
  filter(is.na(Cluster)==F) %>%
  mutate(Cluster_ID=paste("V",Cluster+1,sep=""),
         Cluster=as.character(Cluster),
         Supercluster_ID=gsub("-","_",Supercluster),
         Supercluster_ID=gsub(" ","_",Supercluster_ID))
CT_annot1 <- CT_annot %>% select(Cluster, Supercluster)

dat <- dat %>%
  left_join(CT_annot1, by="Cluster")

# 2. Map to cell tSNE coordinates
dat.tsne <- fread(here("data/neuron_tsne.csv.gz"))
dat <- dat %>%
  inner_join(dat.tsne, by="CellID")


# 3. Get cluster ID for significant clusters of trait (schizophrenia)

dat.cluster.res <- read_xlsx("../data/Supplementary_Datasets.xlsx", sheet = "Supplementary_Data_8") %>%
  filter(Phenotype=="scz2022") %>%
  separate(col=Cluster,into=c("V1","Cluster0"),sep=" \\(", remove=F) %>%
  mutate(Cluster0=as.character(gsub(")","",Cluster0)),
         minuslog10P=-log10(P)) %>%
  select(Cluster0, minuslog10P, if.sig.fdr)

# 4. plot: tSNE of different Superclusters, with labels

#- prepare supercluster labels
dat <- dat %>%
  mutate(Cluster = as.character(Cluster)) 

dat <- dat %>%
  left_join(CT_annot1, by = "Cluster")

# Verify the join worked
table(dat$Supercluster, useNA = "always")

# Now create labels
mylab <- dat %>%
  group_by(Supercluster) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup() %>%
  mutate(
    TSNE1 = ifelse(Supercluster == "Miscellaneous", TSNE1 - 15, TSNE1),
    TSNE2 = ifelse(Supercluster == "Miscellaneous", TSNE2 + 14, TSNE2)
  )
#- prepare for colors
dat1 <- dat %>%
  left_join(dat.cluster.res, by=c("Cluster"="Cluster0")) %>%
  mutate(minuslog10P=ifelse(if.sig.fdr=="yes", minuslog10P,NA))

#- plot
p1 <- ggplot(dat1 %>% mutate(tmp="not sig. (FDR)"),
             aes(x=TSNE1, y=TSNE2, color=minuslog10P, fill=tmp)) +
  geom_point(alpha=.1, size=.1) +
  theme_classic()+
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name="scz2022\n-log10(P)")+
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values=NA, name="") +
  guides(fill=guide_legend(override.aes=list(fill="#E5E5E5",
                                             size=2))) +
  annotate(geom="text", x=mylab$TSNE1, y=mylab$TSNE2, label=mylab$Supercluster,
           color="black") +
  theme(legend.position="bottom", legend.box = "horizontal")

ggsave(p1,
       filename = "../figures/3a_tSNE_overall_scz2022.pdf",
       width = 10, height=10.5)
```

# fig_3b.R
# top 25 clusters for scz2022
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(scico)
library(readxl) 
# read in SCZ cluster level results
df <- read_xlsx("../data/Supplementary_Datasets.xlsx", sheet = "Supplementary_Data_8") |> 
  filter(Phenotype == "scz2022") |> 
  slice_min(P,  n = 25)


p <- ggplot(
  df,
  aes(
    x = reorder(Cluster, -P.fdr),
    y = -log10(P),
    fill = -log10(P)
  )
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("") + ylab(expression('-log'[10]*'(P)')) +
  labs(title="Top Clusters (scz2022)") +
  theme_light() + theme(axis.title.x=element_text(hjust=0)) +
  scico::scale_fill_scico(palette = "lajolla",direction = -1, midpoint = 5) +
  guides(fill="none")

ggsave(p,
  file = "../figures/3b_topClusters_scz.pdf",
  width = 5, height = 6
)

```
My analysis
```{r}
# read in neurotism cluster level results
df_1 <- read_xlsx("../data/Supplementary_Datasets.xlsx", sheet = "Supplementary_Data_8") |> 
  filter(Phenotype == "neuroticism") |> 
  slice_min(P,  n = 25)

p_1 <- ggplot(
  df_1,
  aes(
    x = reorder(Cluster, -P.fdr),
    y = -log10(P),
    fill = -log10(P)
  )
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("") + ylab(expression('-log'[10]*'(P)')) +
  labs(title="Top Clusters (neuroticism)") +
  theme_light() + theme(axis.title.x=element_text(hjust=0)) +
  scico::scale_fill_scico(palette = "lajolla",direction = -1, midpoint = 5) +
  guides(fill="none")

ggsave(p_1,
       file = "../figures/3b_topClusters_neurotism.pdf",
       width = 5, height = 6
)
# read in educattional attaint. cluster level results
df_2 <- read_xlsx("../data/Supplementary_Datasets.xlsx", sheet = "Supplementary_Data_8") |> 
  filter(Phenotype == "educational_attainment") |> 
  slice_min(P,  n = 25)

p_2 <- ggplot(
  df_2,
  aes(
    x = reorder(Cluster, -P.fdr),
    y = -log10(P),
    fill = -log10(P)
  )
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("") + ylab(expression('-log'[10]*'(P)')) +
  labs(title="Top Clusters (edu)") +
  theme_light() + theme(axis.title.x=element_text(hjust=0)) +
  scico::scale_fill_scico(palette = "lajolla",direction = -1, midpoint = 5) +
  guides(fill="none")

ggsave(p_2,
       file = "../figures/3b_topClusters_edu.pdf",
       width = 5, height = 6
)

```

# fig_3a_educational_attainment.Rmd
# analogic as fig3a.R but for educational attainment
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(tidyverse)
library(loomR)
library(here)
library(ggplot2)
library(readxl)

```

```{r}
# 1. LOOM
lfile <- connect(filename = "../data/adult_human_20221007.loom", mode = "r", skip.validate = TRUE)
CellID <- lfile[["col_attrs/CellID"]][]
Cluster <- lfile[["col_attrs/Clusters"]][]
dat <- cbind(CellID, Cluster) %>% as.data.frame()

```

```{r}
# 2. CLuster annotation
CT_annot <- read_excel("../data/silleti_table_s2.xlsx") %>%
  filter(!is.na(Cluster)) %>%
  mutate(Cluster = as.character(Cluster),
         Supercluster = gsub("-| ", "_", Supercluster)) %>%
  dplyr::select(Cluster, Supercluster)

dat <- dat %>%
  mutate(Cluster = as.character(Cluster)) %>%
  left_join(CT_annot, by = "Cluster")

# 3. t-SNEcoordinates
dat.tsne <- fread(here("data/neuron_tsne.csv.gz"))
dat <- dat %>% inner_join(dat.tsne, by = "CellID")
```
```{r}

# 4. data S-LDSC – EDUCATIONAL_ATTAINMENT
dat.cluster.res <- read_xlsx(("../data/Supplementary_Datasets.xlsx"), sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "educational_attainment") %>%
  separate(Cluster, into = c("V1", "Cluster0"), sep = " \\(", remove = FALSE) %>%
  mutate(Cluster0 = gsub("\\)", "", Cluster0),
         minuslog10P = -log10(P)) %>%
  dplyr::select(Cluster0, minuslog10P, if.sig.fdr)

mylab <- dat %>%
  group_by(Supercluster) %>%
  summarise(TSNE1 = mean(TSNE1), TSNE2 = mean(TSNE2)) %>%
  ungroup()


dat1 <- dat %>%
  left_join(dat.cluster.res, by = c("Cluster" = "Cluster0")) %>%
  mutate(minuslog10P = ifelse(if.sig.fdr == "yes", minuslog10P, NA))
```

```{r}
# 5. plot
p_edu <- ggplot(dat1 %>% mutate(tmp = "not sig. (FDR)"),
                  aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = .1, size = .1) +
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla",
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "educational attainment\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") +
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5", size = 2))) +
  annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, label = mylab$Supercluster,
           color = "black") +
  theme(legend.position = "bottom", legend.box = "horizontal")

# 6. Save
ggsave(p_edu,
       filename = ("../figures/3a_tSNE_educational_attainment.pdf"),
       width = 10, height = 10.5)

```
# fig_3a_neuroticism.Rmd
# analogic as fig3a.R but for neuroticism
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(tidyverse)
library(loomR)
library(here)
library(ggplot2)
library(readxl)
```

```{r}
# 1. LOOM
lfile <- connect(filename = "../data/adult_human_20221007.loom", mode = "r", skip.validate = TRUE)
CellID <- lfile[["col_attrs/CellID"]][]
Cluster <- lfile[["col_attrs/Clusters"]][]
dat <- cbind(CellID, Cluster) %>% as.data.frame()

```

```{r}
# 2. CLuster annotation
CT_annot <- read_excel("../data/silleti_table_s2.xlsx") %>%
  filter(!is.na(Cluster)) %>%
  mutate(Cluster = as.character(Cluster),
         Supercluster = gsub("-| ", "_", Supercluster)) %>%
  dplyr::select(Cluster, Supercluster)

dat <- dat %>%
  mutate(Cluster = as.character(Cluster)) %>%
  left_join(CT_annot, by = "Cluster")

# 3. t-SNEcoordinates
dat.tsne <- fread(here("data/neuron_tsne.csv.gz"))
dat <- dat %>% inner_join(dat.tsne, by = "CellID")
```

```{r}

# 4. data S-LDSC – NEUROTICISM
dat.cluster.res <- read_xlsx(("../data/Supplementary_Datasets.xlsx"), sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "neuroticism") %>%
  separate(Cluster, into = c("V1", "Cluster0"), sep = " \\(", remove = FALSE) %>%
  mutate(Cluster0 = gsub("\\)", "", Cluster0),
         minuslog10P = -log10(P)) %>%
  dplyr::select(Cluster0, minuslog10P, if.sig.fdr)

mylab <- dat %>%
  group_by(Supercluster) %>%
  summarise(TSNE1 = mean(TSNE1), TSNE2 = mean(TSNE2)) %>%
  ungroup()


dat1 <- dat %>%
  left_join(dat.cluster.res, by = c("Cluster" = "Cluster0")) %>%
  mutate(minuslog10P = ifelse(if.sig.fdr == "yes", minuslog10P, NA))
```

```{r}
# 5. plot
p_neuro <- ggplot(dat1 %>% mutate(tmp = "not sig. (FDR)"),
                  aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = .1, size = .1) +
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla",
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "neuroticism\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") +
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5", size = 2))) +
  annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, label = mylab$Supercluster,
           color = "black") +
  theme(legend.position = "bottom", legend.box = "horizontal")

# 6. Save
ggsave(p_neuro,
       filename = ("../figures/3a_tSNE_neuroticism.pdf"),
       width = 10, height = 10.5)

```

#category analysis in amygdala clusters
```{r}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(pheatmap)

```

```{r}
df <- read.csv("../data/amygdala_sldsc.csv")
glimpse(df)

```

```{r}
p1 <- ggplot(df, aes(x = factor(Cluster_num), y = Enrichment)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Enrichment in clusters", x = "Clusters", y = "Enrichment") +
  theme_minimal()

p2 <- ggplot(df, aes(x = factor(Cluster_num), y = Prop._h2)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Prop._h2 in clusters", x = "Clusters", y = "Prop._h2") +
  theme_minimal()

ggarrange(p1, p2, ncol = 2, nrow = 2)

```

```{r}
# data prepare _ top 30 enrichment
top_categories <- df %>%
  group_by(Category) %>%
  summarise(var_enrichment = var(Enrichment, na.rm = TRUE)) %>%
  arrange(desc(var_enrichment)) %>%
  slice_head(n = 30) %>%
  pull(Category)

heat_data <- df %>%
  filter(Category %in% top_categories) %>%
  group_by(Category, Cluster_num) %>%
  summarise(mean_enrichment = mean(Enrichment, na.rm = TRUE)) %>%
  pivot_wider(names_from = Cluster_num, values_from = mean_enrichment) %>%
  column_to_rownames("Category")

# Heatmap
pheatmap(
  as.matrix(heat_data),
  scale = "row",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  fontsize_row = 8,     
  fontsize_col = 10,    
  main = "Heatmap average Enrichment (Top 30 categories)"
)

```

#T-sne plot for amygdala excitatory

```{r}
library(data.table)
library(tidyverse)
library(ggplot2)

# Read first 5 rows
sample_data <- fread("../data/amygdala_tsne.csv.gz", nrows = 10)
print("Sample data:")
print(sample_data)

amy_data <- fread("../data/amygdala_tsne.csv.gz") 

# Basic t-SNE plot
plot_amy <- ggplot(amy_data, aes(x = TSNE1, y = TSNE2)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "AMygdala Excitatory t-SNE Projection") +
  theme_minimal()

if ("Cluster" %in% colnames(hipp_data)) {plot_amy <-
  ggplot(amy_data, aes(x = TSNE1, y = TSNE2, color = as.factor(Cluster))) +
    geom_point(alpha = 0.7) +
    scale_color_viridis_d() +
    labs(title = "AMygdala Excitatory t-SNE by Cluster",
         color = "Cluster") +  
    theme_minimal()
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_amygdala_excitatory.pdf")
ggsave(
  plot = plot_amy,
  filename = output_path,
  width = 14,
  height = 10
)

```

#T-sne plot for Hippocampal CA1-3

```{r}
library(tidyverse)
library(ggplot2)

# Read first 5 rows
sample_data <- fread("../data/hipp_tsne.csv.gz", nrows = 10)
print("Sample data:")
print(sample_data)

hipp_data <- fread("../data/hipp_tsne.csv.gz") 

# Basic t-SNE plot
plot_hip <- ggplot(hipp_data, aes(x = TSNE1, y = TSNE2)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "Hippocampal CA1-3 t-SNE Projection") +
  theme_minimal()

if ("Cluster" %in% colnames(hipp_data)) {plot_hip <-
  ggplot(hipp_data, aes(x = TSNE1, y = TSNE2, color = as.factor(Cluster))) +
    geom_point(alpha = 0.7) +
    scale_color_viridis_d() +
    labs(title = "Hippocampal CA1-3 t-SNE by Cluster",
         color = "Cluster") +  
    theme_minimal()
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_hippocampalca13.pdf")
ggsave(
  plot = plot_hip,
  filename = output_path,
  width = 14,
  height = 10)
```

# Analysis for amygdala exc. an hipp. ca1-3
#TSNE amygdala exc. scz2022
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)


# 1. tsne data
dat.tsne <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy")
names(dat.tsne)
# 2. sldsc data
dat.cluster.res.amy <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Amygdala excitatory")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res.amy, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 0.75,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}


output_path <- file.path(getwd(), "..", "figures", "tSNE_amygdala_excitatory_scz2022.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
#SNP-heritability enrichment by cluster- AMygdala exc- data from paper

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep

dat.cluster.res.amy <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Amygdala excitatory")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 2. plot
p <- ggplot(dat.cluster.res.amy, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Amygdala Excitatory",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_amygdala_excitatory_enrichment_sorted.pdf",
       plot = p,
       width = 10, 
       height = 6)

```
#SNP-heritability enrichment by cluster- AMygdala exc- my sldsc data

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep
dat.cluster.res.my <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy\n")
names(dat.tsne)
df <- fread("../data/amygdala_sldsc.csv")

# 3. Function Fisher 
combine_pvalues_fisher <- function(p) {
  chisq_stat <- -2 * sum(log(p), na.rm = TRUE)
  df <- 2 * sum(!is.na(p))
  p_combined <- pchisq(chisq_stat, df = df, lower.tail = FALSE)
  return(p_combined)
}

# 4. Agregation p-value (Fisher)
dat.cluster.res.my <- df %>%
  group_by(Cluster_num) %>%
  summarise(
    combined_pvalue = combine_pvalues_fisher(Enrichment_p),
    minuslog10P = -log10(combined_pvalue),
    if.sig.fdr = ifelse(combined_pvalue < 0.05, "yes", "no")
  ) %>%
  # Sort by minuslog10P in descending order and convert to factor
  arrange(desc(minuslog10P)) %>%
  mutate(Cluster_num = factor(Cluster_num, levels = Cluster_num))

# 2. plot
p <- ggplot(dat.cluster.res.my, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         midpoint = 0.4,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Amygdala Excitatory",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_amygdala_excitatory_enrichment_sorted_my.pdf",
       plot = p,
       width = 10, 
       height = 6)

```
tsne plot scz2022- AMygdala exc- data from paper
```{r}
# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res.amy, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_amygdala_excitatory_scz2022_sup.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
TSNE plot  Fisher combined p-values for AMygdala - my sldsc data
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(scico)

# 1. load data
dat.tsne <- fread("../data/amygdala_tsne.csv.gz")
cat("tsne amy\n")
names(dat.tsne)
df <- fread("../data/amygdala_sldsc.csv")

# 3. Function Fisher 
combine_pvalues_fisher <- function(p) {
  chisq_stat <- -2 * sum(log(p), na.rm = TRUE)
  df <- 2 * sum(!is.na(p))
  p_combined <- pchisq(chisq_stat, df = df, lower.tail = FALSE)
  return(p_combined)
}

# 4. Agregation p-value (Fisher)
dat.cluster.res <- df %>%
  group_by(Cluster_num) %>%
  summarise(
    combined_pvalue = combine_pvalues_fisher(Enrichment_p),
    minuslog10P = -log10(combined_pvalue),
    if.sig.fdr = ifelse(combined_pvalue < 0.05, "yes", "no")
  ) %>%
  mutate(Cluster_num = as.character(Cluster_num))

cat("sldsc data\n")
print(dat.cluster.res)

# 5. Merge  tsne
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%
  left_join(dat.cluster.res, by = "Cluster_num")  

# 6. Labels 
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 7. Plot TSNE
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "-log10(P) (Fisher combined)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5", size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# 8. adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}

# 9. file save
output_path <- file.path(getwd(), "..", "figures", "tsne_amygdala_sldsc_fisher.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)


```
tsne plot for hippocampal ca1-3- data from paper
```{r}
# 1. tsne data
dat.tsne <- fread("../data/hipp_tsne.csv.gz")

dat.cluster.res <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Hippocampal CA1-3")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 
# 3. merge data
dat <- dat.tsne %>%
  mutate(Cluster_num = as.character(Cluster)) %>%  
  left_join(dat.cluster.res, by = "Cluster_num")  
# 4. labels
mylab <- dat %>%
  group_by(Cluster_num) %>%
  summarise(
    TSNE1 = mean(TSNE1),
    TSNE2 = mean(TSNE2)
  ) %>%
  ungroup()

# 5. tsne plot
p1 <- ggplot(dat %>% mutate(tmp = "not sig. (FDR)"),
             aes(x = TSNE1, y = TSNE2, color = minuslog10P, fill = tmp)) +
  geom_point(alpha = 0.5, size = 1) + 
  theme_classic() +
  scico::scale_color_scico(palette = "lajolla", 
                           direction = -1,
                           na.value = "#E5E5E5", midpoint = 5,
                           name = "scz2022\n-log10(P)") +
  xlab("tSNE1") + ylab("tSNE2") + 
  scale_fill_manual(values = NA, name = "") +
  guides(fill = guide_legend(override.aes = list(fill = "#E5E5E5",
                                               size = 2))) +
  theme(legend.position = "bottom", legend.box = "horizontal")

# adding labels
if(nrow(mylab) > 0) {
  p1 <- p1 + 
    annotate(geom = "text", x = mylab$TSNE1, y = mylab$TSNE2, 
             label = mylab$Cluster_num, color = "black")
}
output_path <- file.path(getwd(), "..", "figures", "tSNE_hippocampal_ca1-3_scz2022.pdf")

ggsave(
  plot = p1,
  filename = output_path,
  width = 10,
  height = 10.5
)
```
#SNP-heritability enrichment by cluster- hippocampal ca1-3 - data from paper
```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(stringr)
library(scico) 

# 1. data load and prep
dat.cluster.res.sup <- read_xlsx("../data/Supplementary_Datasets.xlsx", 
                           sheet = "Supplementary_Data_8") %>%
  filter(Phenotype == "scz2022") %>%
  filter(str_detect(Cluster, "Hippocampal CA1-3")) %>%
  mutate(
    Cluster_num = as.numeric(str_extract(Cluster, "(?<=\\()\\d+(?=\\))")),
    minuslog10P = -log10(P),
    is_significant = if.sig.fdr == "yes",
    Cluster_num = factor(Cluster_num) 
  ) %>%
  arrange(desc(minuslog10P)) %>% 
  mutate(Cluster_num = fct_inorder(Cluster_num)) 

# 2. plot
p <- ggplot(dat.cluster.res.sup, 
           aes(x = Cluster_num, 
               y = minuslog10P,
               fill = minuslog10P)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  scico::scale_fill_scico(palette = "lajolla", 
                         direction = -1,
                         name = "-log10(p-value)") +
  labs(
    title = "Schizophrenia (scz2022) - Hippocampal CA1-3",
    subtitle = "SNP-heritability enrichment by cluster (sorted by significance)",
    x = "Cluster number",
    y = "-log10(p-value)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.x = element_blank()
  )

# 3. save
print(p)
ggsave("../figures/scz2022_hippocampal_enrichment_sorted.pdf",
       plot = p,
       width = 10, 
       height = 6)

```

GO analysis
#venn diagram
# 1. Loading gene specificity data
# Assuming the file has columns: 'ENSGID', 'Supercluster_ID', 'specificity'
# and contains information about TDEP genes for each supercluster

```{r}
library(dplyr)
library(readr)
library(VennDiagram)
library(grid)
```
Load data from gzipped TSV file
```{r}
super_df <- tryCatch({
  read_tsv(paste0("../data/supercluster_gene_specificity.tsv.gz"), 
           col_types = cols(
             ENSGID = col_character(),
             Supercluster_ID = col_character(),
             specificity = col_double(),
             .default = col_guess()
           ))
}, error = function(e) {
  stop("Error loading data file: ", e$message)
})
```

```{r}
# Check data structure
message("\nLoaded ", nrow(super_df), " rows from supercluster_gene_specificity.tsv.gz")
message("Available Supercluster_IDs: ", paste(unique(super_df$Supercluster_ID), collapse = ", "))

# Select TDEP genes: top 10% most specific genes in each Supercluster_ID
tdep_genes <- super_df %>%
  group_by(Supercluster_ID) %>%
  arrange(desc(specificity)) %>%
  filter(row_number() <= ceiling(0.10 * n())) %>%
  ungroup()

message("\nSelected ", nrow(tdep_genes), " TDEP genes (top 10% specificity) across all superclusters")

```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)

```

```{r}
# Helper function for GO analysis and dotplot visualization
perform_go_analysis <- function(gene_list, label) {
  if (length(gene_list) < 3) {
    message("Too few genes (", length(gene_list), ") for GO BP analysis in '", label, "'. Skipping.")
    return(NULL)
  }
  
  ego <- tryCatch(
    {
      enrichGO(gene          = gene_list,
               OrgDb         = org.Hs.eg.db,
               keyType       = "ENSEMBL",
               ont           = "BP",
               pAdjustMethod = "BH",
               pvalueCutoff  = 0.05,
               qvalueCutoff  = 0.2)
    },
    error = function(e) {
      message("Error in enrichGO for '", label, "': ", e$message)
      return(NULL)
    }
  )
  
  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
    message("No significant GO BP results for '", label, "'.")
    return(NULL)
  }
  
  message("Found ", nrow(as.data.frame(ego)), " GO BP categories for '", label, "'.")
    
  
  # Create dotplot
  p <- dotplot(ego, showCategory=10) +
    ggtitle(paste("GO BP:", label))
  
  print(p)
  
  # Save the plot
  filename <- paste0("../figures/GO_BP_", gsub(" ", "_", label), ".png")
  ggsave(filename, plot = p, width = 10, height = 6, dpi = 300)
  message("Saved plot as: ", filename)
  
  return(ego)
}
```



```{r}
# Perform GO analysis for unique genes in each supercluster
message("GO BP enrichment analysis for unique genes in 'Amygdala excitatory'")
ego_Amygdala <- perform_go_analysis(unique_Amygdala, "Amygdala excitatory")

message("GO BP enrichment analysis for unique genes in 'Hippocampal CA1-3'")
ego_Hippocampal <- perform_go_analysis(unique_Hippocampal, "Hippocampal CA1-3")
ye
message("GO BP enrichment analysis for common genes")
ego_Common <- perform_go_analysis(common_genes, "Common genes for both superclusters")

```